Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-06-20 manuscript.Rmd
more cuts

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home [@CliffReilly2009; @Barbaro2019; @BruijnsTruelove2020; @LoboHall2019]. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commercially-available and inexpensive IMUs have > 12 hour battery life with onboard storage to record *real-time*, *full-day* motion data at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Indeed, past work has successfully recorded the rate of leg kicks [e.g., @DengTrujillo-Priego2019] and activity intensity [@SchnellerBentsen2017] in infants and children across multiple days.

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity [e.g., @HagerTilton2017; @ArmstrongCovington2019].

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in "tummy time", or they can crawl in a myriad of ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, or in a "W" position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019]. Although @GreenspanCunha2021 were able to use cut-point definitions about infant trunk angle to categorize body positions, this method was not able to differentiate times that infants were held by caregivers off the ground. Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [@Preece2009; @Arif2015], children [@NamPark2013; @RenDing2016; @StewartNarayanan2018], and infants [@YaoPlotz2019; @FranchakScott2021; @AiraksinenRasanen2020]. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest model, that can be later used to predict categories from IMU data that was not annotated. Crucially, this allows automatic classification to scale to full day recording by relying on a relatively smaller set of video annotations.

Three prior investigations have used different machine learning techniques to categorize infant body position from IMUs towards the goal of collecting full-day data. @AiraksinenRasanen2020 tested 4- to 8-month-olds in a laboratory visit, and found 95% accuracy in distinguishing between body position categories that crawling infants could perform on the floor (excluding times that infants were held by caregivers). Using a wider age range of 6-18 months, @FranchakScott2021 found 98% accuracy (*kappa* = 95%) in a laboratory validation study in categorizing body position that included infants who could both crawl and walk and also included a category for caregiver holding. Most recently, @AiraksinenGallen2022 conducted a validation study of body position classification in either a home or clinic testing in 4- to 19-month-olds, refining their previous method to detect moments that infants were carried by caregivers. Classification accuracy did not vary between lab and home settings, and was generally high (95%, *kappa* = .93). Although all three studies yielded promising classification accuracy, accuracy was assessed in brief (15-60 minute) sessions supervised by a researcher, leaving the open question of how well body position classification will scale to testing across an entire day of natural home life.

>>>>

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home [@CliffReilly2009; @Barbaro2019; @BruijnsTruelove2020; @LoboHall2019]. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully *mobile*, and they are *unobtrusive* because they do not require a researcher to follow with a camcorder. Many commercially-available IMUs have > 12 hour battery life with onboard storage to record *real-time*, *full-day* motion data at a high sampling rate (e.g., 50-100 Hz).

The open question is whether *automatic classification* is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is a complex because movement can vary greatly *within* a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in "tummy time", or they can crawl in a myriad of ways [@AdolphVereijken1998]. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a "V" position with legs fully extended, or in a "W" position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity [@KwonZavos2019; @PatelShi2019].

Modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults [@Preece2009; @Arif2015], children [@NamPark2013; @RenDing2016; @StewartNarayanan2018], and infants [@YaoPlotz2019; @FranchakScott2021; @AiraksinenRasanen2020]. Three prior investigations have used different machine learning techniques to categorize infant body position from IMUs towards the goal of collecting full-day data. @AiraksinenRasanen2020 tested 4- to 8-month-olds in a laboratory visit, and found 95% accuracy in distinguishing between body position categories that crawling infants could perform on the floor (excluding times that infants were held by caregivers). Using a wider age range of 6-18 months, @FranchakScott2021 found 98% accuracy (*kappa* = 95%) in a laboratory validation study in categorizing body position that included infants who could both crawl and walk and also included a category for caregiver holding. Most recently, @AiraksinenGallen2022 conducted a validation study of body position classification in either a home or clinic testing in 4- to 19-month-olds, refining their previous method to detect moments that infants were carried by caregivers. Classification accuracy did not vary between lab and home settings, and was generally high (95%, *kappa* = .93). Although all three studies yielded promising classification accuracy, accuracy was assessed in brief (15-60 minute) sessions supervised by a researcher, leaving the open question of how well body position classification will scale to testing across an entire day of natural home life.

### 2023-06-20 manuscript.tex
more cuts

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home \autocite{CliffReilly2009,Barbaro2019,BruijnsTruelove2020,LoboHall2019}. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully \emph{mobile}, and they are \emph{unobtrusive} because they do not require a researcher's presence nor do they record sensitive audio/video that might influence participants' behavior. Many commercially-available and inexpensive IMUs have \textgreater{} 12 hour battery life with onboard storage to record \emph{real-time}, \emph{full-day} motion data at a sampling rate that is higher than typical video (e.g., 50-100 Hz). Indeed, past work has successfully recorded the rate of leg kicks \autocite[e.g.,][]{DengTrujillo-Priego2019} and activity intensity \autocite{SchnellerBentsen2017} in infants and children across multiple days.

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. IMUs typically contain accelerometers that measure linear acceleration paired with gyroscopes that measure angular acceleration. Unlike motion tracking systems that might be used in a lab, IMUs do not provide data about the position of the body in space. Thus, data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories (e.g., supine, prone, sitting, upright). The difficulty of the classification task depends on the categories of interest. More basic aspects of movement, such as overall activity intensity, can be identified by taking the magnitude of acceleration (irrespective of direction) and applying thresholds or cut-points to define when a movement has occurred at a particular intensity \autocites[e.g.,][]{HagerTilton2017,ArmstrongCovington2019}.

Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is too complex for cut-point definitions to be accurate. First, the magnitude of movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in ``tummy time'', or they can crawl in a myriad of ways \autocite{AdolphVereijken1998}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, or in a ``W'' position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Indeed, creating an all-encompassing set of rules for how to annotate sitting from video is no trivial task because of the various ways that sitting can occur in daily life. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity \autocite{KwonZavos2019,PatelShi2019}. Although \textcite{GreenspanCunha2021} were able to use cut-point definitions about infant trunk angle to categorize body positions, this method was not able to differentiate times that infants were held by caregivers off the ground. Thus, modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults \autocite{Preece2009,Arif2015}, children \autocite{NamPark2013,RenDing2016,StewartNarayanan2018}, and infants \autocite{YaoPlotz2019,FranchakScott2021,AiraksinenRasanen2020}. Synchronized video with ground-truth human annotations creates training data for a machine learning algorithm, such as a random forest model, that can be later used to predict categories from IMU data that was not annotated. Crucially, this allows automatic classification to scale to full day recording by relying on a relatively smaller set of video annotations.

Three prior investigations have used different machine learning techniques to categorize infant body position from IMUs towards the goal of collecting full-day data. \textcite{AiraksinenRasanen2020} tested 4- to 8-month-olds in a laboratory visit, and found 95\% accuracy in distinguishing between body position categories that crawling infants could perform on the floor (excluding times that infants were held by caregivers). Using a wider age range of 6-18 months, \textcite{FranchakScott2021} found 98\% accuracy (\emph{kappa} = 95\%) in a laboratory validation study in categorizing body position that included infants who could both crawl and walk and also included a category for caregiver holding. Most recently, \textcite{AiraksinenGallen2022} conducted a validation study of body position classification in either a home or clinic testing in 4- to 19-month-olds, refining their previous method to detect moments that infants were carried by caregivers. Classification accuracy did not vary between lab and home settings, and was generally high (95\%, \emph{kappa} = .93). Although all three studies yielded promising classification accuracy, accuracy was assessed in brief (15-60 minute) sessions supervised by a researcher, leaving the open question of how well body position classification will scale to testing across an entire day of natural home life.

>>>>

Measuring infant movement with inertial movement units (IMUs) is a promising avenue for long form recordings of motor behavior in the home \autocite{CliffReilly2009,Barbaro2019,BruijnsTruelove2020,LoboHall2019}. Lightweight sensors (10-30 g) can be embedded in garments to make recordings fully \emph{mobile}, and they are \emph{unobtrusive} because they do not require a researcher to follow with a camcorder. Many commercially-available IMUs have \textgreater{} 12 hour battery life with onboard storage to record \emph{real-time}, \emph{full-day} motion data at a high sampling rate (e.g., 50-100 Hz).

The open question is whether \emph{automatic classification} is sufficiently accurate to measure movement categories that are relevant to developmental and clinical research, and whether measurements continue to be accurate over long recording periods. Data processing algorithms are needed to classify the raw sensor data (i.e., linear and angular acceleration time series) into meaningful categories. Categorizing body position---supine on the back, prone on the belly, sitting, upright, or held off the ground by a caregiver---is a complex because movement can vary greatly \emph{within} a body position. An upright infant can be standing still or can be walking briskly across the room. A prone infant can be stationary in ``tummy time'', or they can crawl in a myriad of ways \autocite{AdolphVereijken1998}. Moreover, the configuration of the arms, legs, and torso within a body position can vary greatly in everyday contexts. Infants can sit on the floor in a tripod position with support from an arm, in a ``V'' position with legs fully extended, or in a ``W'' position with knees bent. Sitting on a caregiver's lap without the need to maintain balance means that the legs can dangle and the torso can lean in different directions. Sitting in a high chair or car seat reduces the magnitude of postural sway within sitting, and creates even more possibilities for how the arms and legs may move relative to the torso. Finally, caregivers frequently pick up and transport infants, creating motion signals that need to be differentiated from independent activity \autocite{KwonZavos2019,PatelShi2019}.

Modern approaches to human activity recognition have used machine learning to classify activity categories based on features derived from IMU data in adults \autocite{Preece2009,Arif2015}, children \autocite{NamPark2013,RenDing2016,StewartNarayanan2018}, and infants \autocite{YaoPlotz2019,FranchakScott2021,AiraksinenRasanen2020}. Three prior investigations have used different machine learning techniques to categorize infant body position from IMUs towards the goal of collecting full-day data. \textcite{AiraksinenRasanen2020} tested 4- to 8-month-olds in a laboratory visit, and found 95\% accuracy in distinguishing between body position categories that crawling infants could perform on the floor (excluding times that infants were held by caregivers). Using a wider age range of 6-18 months, \textcite{FranchakScott2021} found 98\% accuracy (\emph{kappa} = 95\%) in a laboratory validation study in categorizing body position that included infants who could both crawl and walk and also included a category for caregiver holding. Most recently, \textcite{AiraksinenGallen2022} conducted a validation study of body position classification in either a home or clinic testing in 4- to 19-month-olds, refining their previous method to detect moments that infants were carried by caregivers. Classification accuracy did not vary between lab and home settings, and was generally high (95\%, \emph{kappa} = .93). Although all three studies yielded promising classification accuracy, accuracy was assessed in brief (15-60 minute) sessions supervised by a researcher, leaving the open question of how well body position classification will scale to testing across an entire day of natural home life.
